Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-20 analysis/paper/paper.Rmd
edited and fixed typos in the result section

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
To assess MCMC analysis, we first evaluate three diagnostics presented as visual summaries, including density plots, trace plots, and autocorrelation plots [@Hamra2013]. Those plots provide an informal way to diagnose model convergence by identifying any unexpected peaks or shapes in distribution. Then we access output from our two  models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. We set of 10000 network graphs simulated from the estimated posterior distribution in ERGMs. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network.

@@

<!--posterior stats for pre-E-->
We compared estimated statistics from the posterior distributions, networks before and after the European presence, to examine their differences in network structure (Table XX). For the former network, the nodal covariates, age-homophily and ritual-homophily, have significant effects on the relations between burials. The age homophiligy shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite other covariates are positive, such as sex-homophily and burial value-homophily, they do not show a significant tendency to forming relations duo to value of zero in a confidence interval. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly. For the endogenous network effects, both gwesp and gwdegree present significant positive effects. The positive value of gwesp suggests a tendency of burials to be clustered in closed transitive structures, while positive gwdegree indicates a tendency toward centralization that means the presence of a limited number of burials with more relations.

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans, there are no significant effects of nodal and dyadic covariates, but we still have a clue of possible tendency. The nodal covariates of age-homophily, sex-homophily, ritual-homophily, and dyadic covariate of physical distance all show positive effects, while the burial value-homophily has a negative effect. The interesting thing is that the burials in the same wealth level may tend not to form relations. Similar to the network before the European arrival, the endogenous network variables, gwesp and gwdegree, all demonstrate positive effects significantly. However, there are differences in values of positive effects between two networks. The later network has a smaller gwesp effect and a larger gwedegree effect that may suggest a less tendency toward clustering but more tendency toward centralization after the European presence. Both posterior estimates presents symmetric distributions because of the posterior means are close to posterior medians (Table XX).

@@

<!--assessment of MCMC analysis and BGOF--> 
Bayesian goodness of fit (GOF) diagnostics plots show that both models fit the observed networks very well for the minimum geodesic distance distribution (i.e. the number of edges between node pairs in a shortest path) and the degree distribution. For edgewise shared partner (gwesp) distribution, despite some observations falls outside the 95% interval, the fitting is generally good with most observations within it. This also suggests that our observed network statistic does not follow a normal distribution, especially for the network after the arrival of the Europeans that demonstrate a bimodal pattern. We also compared the first three distribution moments of each observed distributions and their corresponding simulated distributions represented by mean. The figure shows that modeled values from pre-Euroepan network are closer to the observed values overall, compared to the values from the post-European network. For the variance of degree, the values from modeled post-European distribution are closer to the observed distribution. In general, this demonstrates a better fit for the pre-European network. However, we should note that the differences between the statistics of the two models are relatively small, which allows direct comparison of two networks.

>>>>

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
To assess MCMC analysis, we first evaluated three diagnostics visual summaries, including density plots, trace plots, and autocorrelation plots [@Hamra2013]. Those plots provide an informal way to diagnose model convergence by identifying any unexpected peaks or shapes in distribution. Then we access output from our two  models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. We set of 10000 network graphs simulated from the estimated posterior distribution in ERGMs. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network.

@@

<!--posterior stats for pre-E-->
We compared estimated statistics from the posterior distributions, networks before and after the European presence, to examine their differences in network structure (Table XX). For the former network, the nodal covariates, age-homophily and ritual-homophily, have significant effects on the relations between burials. The age homophiligy shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive value from some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to forming relations duo to value of zero in a confidence interval. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly. For the endogenous network effects, both gwesp and gwdegree present significant positive effects. The positive value of gwesp suggests a tendency of burials to be clustered in closed transitive structures, while positive gwdegree indicates a tendency toward centralization that means the presence of a limited number of burials with more relations.

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans, there are no significant effects of nodal and dyadic covariates, but we still have a clue of possible tendency. The nodal covariates of age-homophily, sex-homophily, ritual-homophily, and dyadic covariate of physical distance all show positive effects, while the burial value-homophily has a negative effect. We also found that the burials in the same wealth level may tend not to form relations. Similar to the earlier network, the endogenous network variables, gwesp and gwdegree, all demonstrate positive effects significantly. However, there are differences in values of positive effects between two networks. The later network has a smaller gwesp effect and a larger gwedegree effect that may suggest a less tendency toward clustering but more tendency toward centralization after the European presence. Both posterior estimates presents symmetric distributions because of the posterior means are close to posterior medians (Table XX).

@@

<!--assessment of MCMC analysis and BGOF--> 
Bayesian goodness of fit (GOF) diagnostics plots show that both models fit the observed networks very well for the minimum geodesic distance distribution (i.e. the number of edges between node pairs in a shortest path [@Hunter2008]) and the degree distribution. For edgewise shared partner (gwesp) distribution, despite some observations falls outside the 95% interval, the fitting is generally good with most observations within it. This also suggests that our observed network statistics do not follow a normal distribution, especially for the network after the arrival of the Europeans that presents a bimodal pattern. We also compared the first three distribution moments of each observed distribution and their corresponding simulated distribution represented by mean. The figure shows that modeled values from the pre-European network are closer to the observed values overall, compared to the values from the post-European network. For the variance of degree, the values from modeled post-European distribution are closer to the observed distribution. In general, this demonstrates a better fit for the pre-European network. However, it should be noted that the differences between the statistics of two models are relatively small, which still allows direct comparison of two networks.
